Computational methods in landscape ecology

### 2024-10-16 Review.qmd
Updates review

---
title: "Reviewer Comments"
format:
  pdf:
    keep-tex: false
    include-in-header:
      text: |
        \setlength\parindent{0pt}

@@

\textcolor[HTML]{0BB67D}{Reply: Thank you very much for the encouraging words. We will use the valuable and helpful feedback to improve the manuscript.}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply: We added the term model-driven as suggested.}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply: We agree that landscape metrics can be extracted straightforwardly from raster data, and we are adding this as another advantage of using them.}

@@

\textcolor[HTML]{0BB67D}{Reply: Indeed we forgot to add a subsection title which is added now.}

@@

\textcolor[HTML]{0BB67D}{Reply: Added as suggested.}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply: Similar to landscape metrics also surface metrics are scale-dependent because both are designed to quantify landscape heterogeneity - which is inherently scale-dependent. We added this explanation including an according literature references.}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

@@

\textcolor[HTML]{0BB67D}{Reply:}

>>>>

---
title: ""
format:
  pdf:
    keep-tex: false
    include-in-header:
      text: |
        \setlength\parindent{0pt}

@@

\textcolor[HTML]{007895}{Reply: Thank you very much for the encouraging words. We will use the valuable and helpful feedback to improve the manuscript.}

@@

@@

\textcolor[HTML]{007895}{Reply: This is a very helpful comment about the general structure of the manuscript and we tried to address it as much as possible. This included several points. First, we agree that a general section on data is useful for readers. Thus, we added such a section at the beginning of the manuscript. Second, we modified especially the first paragraphs of the "Spatial patterns" (sub-)sections in order to improve the logical flow of the manuscript. Third, we rearranged and modified the "Connectivity" section as well. We removed the "Structural connectivity" sub-section by including the key statements into the general paragraph about connectivity.}

@@

\textcolor[HTML]{007895}{Reply: We added the term "model-driven" as suggested.}

@@

\textcolor[HTML]{007895}{Reply: Indeed, such a table is very helpful. Thus, we removed all paragraphs that only list software from the specific sections and now list all software in a central table (Tab. 1).}

@@

\textcolor[HTML]{007895}{Reply: We agree that landscape metrics can be extracted straightforwardly from raster data, and we are adding this as an advantage related to them.}

@@

\textcolor[HTML]{007895}{Reply: Indeed, we forgot to add a sub-section title which is added now.}

@@

\textcolor[HTML]{007895}{Reply: We replaced "entropy" with "measure".}

@@

\textcolor[HTML]{007895}{Reply: We added a brief definition of edge length in this context.}

@@

\textcolor[HTML]{007895}{Reply: Similar to landscape metrics, also surface metrics are scale-dependent because both are designed to quantify landscape heterogeneity - which is inherently scale-dependent. We added this explanation including an according literature reference.}

@@

\textcolor[HTML]{007895}{Reply: We removed the term "query" from the sentence because we agree that it is a source of confusion. However, we believe that the sentence using only "areas" instead is clear enough for readers to follow.}

@@

\textcolor[HTML]{007895}{Reply: We agree that the whole section was confusing for readers unfamiliar with the topic. Thus, we reworded it to make it easier to follow by removing a lot of the database jargon.}

@@

\textcolor[HTML]{e93c26}{Reply: TODO Thanks for pointing this out. We added it to the figure.}

@@

\textcolor[HTML]{e93c26}{Reply: TODO}

@@

\textcolor[HTML]{007895}{Reply: We added "flow" as suggested.}

@@

\textcolor[HTML]{007895}{Reply: We added "dispersal" as suggested.}

@@

\textcolor[HTML]{007895}{Reply: We added dispersal as suggested.}

@@

\textcolor[HTML]{007895}{Reply: We added the specification that movement also includes dispersal of plants. Furthermore, we believe that this aspect is now also highlighted due to the other suggestions of the reviewer that we implemented in this section. Generally, we added this each time movement is mentioned in the section.}

@@

\textcolor[HTML]{007895}{Reply: We added "disperse" as suggested.}

@@

\textcolor[HTML]{007895}{Reply: We are unsure why the point about edge artifacts was added to the manuscript. Thus, we completely removed it here since we think it adds more confusion than helping the readers.}

@@

\textcolor[HTML]{007895}{Reply: We agree that the original statement about underlying assumptions was unclear and reworded the sentence.}

@@

\textcolor[HTML]{007895}{Reply: We added this specification as suggested.}

@@

\textcolor[HTML]{007895}{Reply: We agree that this statement is true for most if not all methods presented in this manuscript. However, we believe it is also helpful to keep it in this section specifically. However, we added a new short "Data" section at the beginning of the manuscript in which we also highlight importance of input data specifically.}

@@

\textcolor[HTML]{007895}{Reply: Thank you very much for adding this insightful remark about the references. We reworded the sentence to ensure the remarks are reflected in the manuscript.}

@@

\textcolor[HTML]{007895}{Reply: We added that digital twins benefit from the increasing rate of data generation in ecology.}

@@

\textcolor[HTML]{007895}{Reply: We renamed the section as suggested.}

@@

\textcolor[HTML]{007895}{Reply: We reworded the sentence being more specific about the corresponding references.}

@@

\textcolor[HTML]{007895}{Reply: Thank you for pointing out this detail. We added this as another potentail drawback of the method.}

@@

\textcolor[HTML]{007895}{Reply: We reworded the sentence stating that there are differences between solvers and that the optimal solver depends on the specific problem. However, the best solver could potentially not be accessible due to software licenses.}

### 2024-10-16 Sections/01-introduction.qmd
Updates review

Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include include but are not limited to pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, including human activities [@Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
<!-- R1: Data-driven and model-driven -->
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven and model-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape for a review of landscape ecology experiments].
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

<!-- R1: Very good idea to suggest software implementing methodological developments. It would be even better with a recapitulative table -->
Here, we introduce recent computational developments in some important topics of landscape ecology (Fig. 1) including software that implements or is potentially capable of implementing them (Tab. 1).
In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
For a more comprehensive and general introduction to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.

>>>>

Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land use systems [@Forman1986LandscapeEcology] with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include but are not limited to pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
<!-- R1: Data-driven and model-driven -->
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven and model-driven approaches [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape for a review of experimental studies].
Additionally, because landscape ecology is a cross-disciplinary field [@With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

<!-- R1: Very good idea to suggest software implementing methodological developments. It would be even better with a recapitulative table -->
Here, we introduce recent computational developments in some important topics of landscape ecology (Fig. 1) including software that implements or is potentially capable of implementing them (Tab. 1).
In this context, we are focusing on open-source software and scripting languages such as R, Python, and Julia.
However, we do not provide a systematic literature review or a general introduction to (computational) landscape ecology.
For such, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.

<!-- R1: The links between Spatial sampling, Spatial planning, and Genetics, towards Landscape mean?  The figure would be more meaningful with an explanation of these links, either in the caption or directly on the figure. Also, for other topics (e.g. connectivity), a keyword (or short sentence) summarizing the contribution of the topic to Landscape Ecology would be great. -->

### 2024-10-16 Sections/02-data.qmd
Updates review

# Data {#sec-data}

@@

There are two fundamental data models used to represents landscapes, namely the raster data and the vector data model. 
Raster data uses regularly spaced grid cells, while vector data uses points, lines, and polygons to represent landscape features [@With2019EssentialsLandscape].
The issue of scale is closely related to the used data model. 
This includes the extent of the study area and resolution of the data (i.e., the smallest data unit).
Additionally, thematic resolution in landscape ecology often refers to the values describing landscape features, e.g., the number of discrete LULC categories.
Various classification systems are relevant at local, regional, and global extent, such as the National Land Cover Database (NLCD), the Coordination of Information on the Environment (CORINE), or the Food and Agriculture Organization (FAO) systems [@Nedd2021SynthesisLand].
Finally, also temporal scales are relevant both in terms of extent and resolution.

Less frequently discussed, but often critical, is the spatial reference system of the data. 
Cartesian coordinate reference systems measure distances between two points in Euclidean distances, typically in meters. 
In contrast, geographic coordinate reference systems are based on degrees and measure distances as great-circle distances.
Numerous coordinate reference systems exist, each with distinct properties that may distort areas, distances, angles, or all three. 
Therefore, it is essential to choose the appropriate spatial reference system based on the available data and the specific research question

Data quality is generally variable, and data products are not flawless and can contain errors and biases.
For example, global land cover products typically have an overall accuracy of 70 to 80%, indicating that 20 to 30% of the grid cells may be misclassified.
Furthermore, these misclassification is not random, but are often correlated with specific LULC types, regions, and seasonal variations [@Liu2021FinerresolutionMapping].
Despite this, many landscape ecology studies tend to accept the data at face value, neglecting its accuracy and inherent uncertainty.

>>>>

# Data models {#sec-data}

@@

There are two fundamental data models used to represents landscapes, namely the raster data and the vector data model. 
Raster data uses regularly spaced grid cells, while vector data uses points, lines, and polygons to represent landscape features [@With2019EssentialsLandscape].
Often, the choice of the data model is driven by data and software availability or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
For example, many LULC maps are provided as raster data as they often relate to underlying remote sensing products.

The issue of scale is closely related to the used data model. 
This includes the extent of the study area and resolution of the data (i.e., the smallest data unit).
Additionally, thematic resolution in landscape ecology often refers to the values describing landscape features, e.g., the number of discrete LULC categories.
Various classification systems are relevant at local, regional, and global extent, such as the National Land Cover Database (NLCD), the Coordination of Information on the Environment (CORINE), or the Food and Agriculture Organization (FAO) systems [@Nedd2021SynthesisLand].
Finally, also temporal scales are relevant both in terms of extent and resolution.

Less frequently discussed, but often critical, is the spatial reference system of the data. 
Cartesian coordinate reference systems measure distances between two points in Euclidean distances, typically in meters. 
In contrast, geographic coordinate reference systems are based on degrees and measure distances as great-circle distances.
Numerous coordinate reference systems exist, each with distinct properties that may distort areas, distances, angles, or all three. 
Therefore, it is essential to choose the appropriate spatial reference system based on the available data and the specific research question.

Data quality is generally variable, and data products are not flawless and can contain errors and biases.
For example, global land cover products typically have an overall accuracy of 70 to 80%, indicating that 20 to 30% of the grid cells may be misclassified.
Furthermore, these misclassifications are not random, but are often correlated with specific LULC types, regions, and seasonal variations [@Liu2021FinerresolutionMapping].
Despite this, many landscape ecology studies tend to accept the data at face value, neglecting its accuracy and inherent uncertainty.

### 2024-10-16 Sections/03-spatial-patterns.qmd
Updates review

Spatial patterns can be defined as the scale-dependent predictability of the physical arrangement of observations [@Dale1999SpatialPattern], or as clearly identifiable structures in nature itself or data extracted from nature [@Grimm1996PatternorientedModelling].
Importantly, patterns observed in nature contain information about the history of the system, such as demographic processes, dispersal characteristics, or climatic patterns [@Wiegand2003UsingPatternoriented].
However, spatial patterns are not only a result of processes, but could also be drivers of them.
Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner2015LandscapeEcology].

@@

Traditionally and at present, most of the prominent approaches that quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
<!-- R1: They are also naturally extracted from remote sensing products (raster data), which often makes them easier to monitor over time -->
The strength of such landscape metrics is that they are easy to apply, communicate, and can be extracted straightforwardly from raster data [e.g., remote sensing products\; @Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
However, also limitations have been identified. 
These limitations include shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

To address some limitations, for example correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees have been used to identify core metrics that capture main components of landscape patterns.
More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
These components of landscape patterns are in line with recent reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] that have similarly suggested that there are also two fundamental components of landscape patterns, namely amount and adjacency.
These components are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory, and later formalized in @Nowosad2019InformationTheory.

@@

The landscape mosaic method offers a way to quantify spatial patterns through a tri-polar classification model involving three LULC classes [@Riitters2009IndicatorForest]. 
The approach uses a moving window to determine the proportions of these three classes within each central grid cell.
These proportions are then classified into 19 mosaic classes based on thresholds that describe the presence, dominance, or uniqueness of each class.
This allows the assessment of content, context, and interface zones of LULC data in a scale-dependent manner.
Recently, the method has been improved by expanding the classification to 103 classes and incorporating heatmap visualizations and summaries [@Vogt2024RevisitingLandscape].

@@

Entropy measures in landscape ecology are mainly derived from information theory and thermodynamics.
They are mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity), as well as scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
<!-- R1: of the selected entropy measure -->
However, studies show that insights gained from entropy measures depend on the formulation of the selected entropy measure, and on the underlining data model, e.g., the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].

@@

Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, there has been a surge in the development of entropy-based metrics for landscape ecology in the last few years.
In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate [the configuration of the system elements\; @Gao2019ComputationBoltzmann].
<!-- R1: the edge length -> Could you specify a bit -->
@Cushman2016CalculatingConfigurational proposed to relate the edge length (defined as the length of neighboring cells with different LULC classes) to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
This approach was later generalized for calculations based on the raster surface model and point patterns [@Cushman2021EntropyLandscape].
Subsequently, @Gao2017HierarchybasedSolution proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
However, the relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion [e.g., @Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial].

Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies, which are both generalizations of the Shannon entropy, have been applied to quantify the landscape complexity.
The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos] as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
Thus, it can be useful in cases where the dissimilarities between LULC classes are relevant.
Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].

@@

Surface metrics are based on the gradient surface model using gridded raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm]. 
These metrics were mostly adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
<!-- R1: I guess that such metrics are also very scale dependent -->
The gradient surface model can increase the resemblance of the data to the natural world because it allows for the inclusion of more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows for the exploration of different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
Nevertheless, similar to landscape metrics, surface metrics are also scale-dependent because they are used to quantify landscape heterogeneity, which is itself scale-dependent [@Frazier2016SurfaceMetrics].
Yet, software to calculate these metrics are still rare and further research is needed into the specific pattern-process links and ecologically meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values [@Riitters2023MappingLandscape].

@@

Many landscape ecology studies are based on the raster data model.
However, the choice of the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
For example, many LULC maps are provided as raster data as they often relate to underlying remote sensing products
However, nowadays, vector data has become increased in availability and has been widely used in other fields, such as geography or urban studies.

Hundreds of metrics have already been developed for raster data, and the most straightforward approach may therefore be to reimplement these same metrics for vector data.
Although this is possible for most of the metrics, there are a few exceptions where metrics are specifically related to the raster data model [@Yao2022VecLIFramework].
For example, in urban planning, vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
Earlier approaches to quantify shape complexity included four categories of compactness measures, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness].

However, also vector-based metrics also have limitations.
The most important limitations relate to computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
Another technical issue is the requirement of topologically correct data (e.g., geometries cannot overlap), which is often especially problematic for data from different sources.
Last, the pattern-process link for vector-based metrics remains underexplored compared to raster-based metrics.

@@

Spatial patterns can be analyzed through a range of computational operations, e.g., comparing, searching, or grouping.
These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures), and can be calculated for different areas or for the same area at two different moments in time.

>>>>

Spatial patterns can be defined as the scale-dependent predictability of the physical arrangement of observations [@Dale1999SpatialPattern] or as clearly identifiable structures in nature itself or data extracted from nature [@Grimm1996PatternorientedModelling].
Importantly, patterns observed in nature contain information about the history of the system, such as demographic processes, dispersal characteristics, or climatic patterns [@Wiegand2003UsingPatternoriented].
However, spatial patterns are not only a result of processes, but could also be drivers of them.
Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner2015LandscapeEcology].

@@

Traditionally and at present, most of the prominent approaches that quantify spatial patterns revolve around raster data using categorical values based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
<!-- R1: They are also naturally extracted from remote sensing products (raster data), which often makes them easier to monitor over time -->
The strength of such landscape metrics is that they are easy to apply, communicate, and can be extracted straightforwardly from raster data based on remote sensing products [@Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
However, also limitations have been identified. 
These limitations include shortcomings in their ability to quantify the spatial structure, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [@Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

To address  limitations related to correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees have been used to identify core metrics that capture main components of landscape patterns.
More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
These components of landscape patterns are in line with recent reviews [@Gustafson2019HowHas; @Riitters2019PatternMetrics] that have similarly suggested that there are two fundamental components of landscape patterns, namely amount and adjacency.
These components are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory, and later formalized in @Nowosad2019InformationTheory.

@@

The landscape mosaic method offers a way to quantify spatial patterns through a tri-polar classification model involving three LULC classes [@Riitters2009IndicatorForest]. 
The approach uses a moving window to determine the proportions of these three classes within each focal grid cell.
These proportions are then classified into 19 mosaic classes based on thresholds that describe the presence, dominance, or uniqueness of each class.
This allows the assessment of content, context, and interface zones of LULC data in a scale-dependent manner.
Recently, the method has been improved by expanding the classification to 103 classes and incorporating heatmap visualizations and summaries [@Vogt2024RevisitingLandscape].

@@

Entropy measures in landscape ecology are mainly derived from information theory and thermodynamics.
They are mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity), as well as scale dependence (spatio-temporal heterogeneity) [@Vranken2015ReviewUse].
<!-- R1: of the selected entropy measure -->
However, studies show that insights gained from entropy measures depend on the formulation of the selected measure, and on the underlining data model, e.g., the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].

@@

Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, there has been a surge in the development of entropy-based metrics for landscape ecology in the last few years.
In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate (the configuration of the system elements) [@Gao2019ComputationBoltzmann].
<!-- R1: the edge length -> Could you specify a bit -->
@Cushman2016CalculatingConfigurational proposed to relate the edge length (defined as the side lengths of neighboring cells with different LULC classes) to the microstate of the landscape and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
This approach was later generalized for calculations based on the raster surface model and point patterns [@Cushman2021EntropyLandscape].
Subsequently, @Gao2017HierarchybasedSolution proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
However, the relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion [@Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial].

Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies, which are both generalizations of the Shannon entropy, have been applied to quantify the landscape complexity.
The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos] as it measures not only the relative abundances of elements but also the pairwise dissimilarities or distances between them.
Thus, it can be useful in cases where the dissimilarities between LULC classes are relevant.
Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy) which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].

@@

Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm]. 
These metrics were mostly adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
<!-- R1: I guess that such metrics are also very scale dependent -->
The gradient surface model can increase the resemblance of the data to the natural world because it allows for the inclusion of more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows for the exploration of different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
Nevertheless, similar to landscape metrics, surface metrics are also scale-dependent because they are used to quantify landscape heterogeneity which is itself scale-dependent [@Frazier2016SurfaceMetrics].
Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecologically meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
Yet, another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values [@Riitters2023MappingLandscape].

@@

<!-- R1: As an example, I found the subsection "Vector metrics" a bit confusing, mainly because, as you say in L159-160, most landscape metrics can be computed on both raster and vector data. -->

Hundreds of metrics have already been developed for raster data, and the most straightforward approach may therefore be to reimplement these same metrics for vector data.
Although this is possible for most of the metrics, there are a few exceptions where metrics are specifically related to the raster data model [@Yao2022VecLIFramework].
For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
Earlier approaches to quantify shape complexity included four categories of compactness measures, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness].

However, also vector-based metrics have limitations.
The most important limitations relate to computational complexity which makes calculations of vector metrics slower than their raster equivalents.
Another technical issue is the requirement of topologically correct data (e.g., geometries cannot overlap) which is often especially problematic for data from different sources.
Last, the pattern-process link for vector-based metrics remains underexplored compared to raster-based metrics.

@@

Spatial patterns can be analyzed through a range of computational operations, e.g., comparing, searching, or grouping.
These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures) and can be calculated for different areas or for the same area at two different moments in time.
